---
title: 'GLM 2: count data'
author: "Timothee Bonnet"
date: "20 May 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Today:

* Fit generalized linear models for count data
* Predict and visualise model output
* Understand issues of over-dispersion and how to address them

Packages:
```{r}
library(ggplot2)
library(performance)
library(DHARMa)
library(glmmTMB)
```



## Count data properties

Ideal count data produced by "Poisson" process. The Poisson distribution has a single parameter: the expected count, lambda.
```{r}
randomcount <- rpois(n = 10000, lambda = 3.2)

mean(randomcount)
var(randomcount)
range(randomcount)

summary(randomcount)

randomcount <- rpois(n = 10000, lambda = 10.8)

mean(randomcount)
var(randomcount)
range(randomcount)

summary(randomcount)

ggplot(data.frame(x=randomcount), aes(x=x))+
  geom_histogram(binwidth = 0.5,aes(y=stat(count) / sum(count)))
```

The mean of a true Poisson variable equals its variance. 
More in general in count variables that are not perfect Poisson, the variance is a function of the mean.

```{r}
x <- seq(from = 1, to=10, by=1)

dat <- data.frame(x= rep(x, each=1000),
                  y=as.vector(sapply(x, function(x) {rpois(n = 1000, lambda = x)})))

ggplot(dat, aes(x=as.factor(x), y=y)) + geom_boxplot()
```


## Fitting count data GLMs in R

The three components of a glm:

1. A linear model: response = intercept + predictor*slope ...
2. A link function between the scale of the data (positive intergers) and the scale of the linear model (-inf to +inf). For count data generally log().
3. A random distribution linking expected data (on average 1.27) to actual data (0,1,2,3...). For count data, the Poisson distribution or other related distributions.




The simplest solution is:

```{r, eval=FALSE}
glm(obs ~ 1 + x , family = "poisson", data=data)
```

**However you should NEVER** use this.
glm( family="poisson") assumes there is no unexplained variation apart from the incompressible Poisson random process, that is, there is no unexplained variation in expected counts, or $V(\mathrm{exp}(Y)) = E(\mathrm{exp}(Y))$.
This is almost never the case with biological data. Most of the time there is lots of unexplained variance and are over-dispersed. On the other hand, some biological processes are "conservative"/under-dispersed and have $V(\mathrm{exp}(y)) < E(\mathrm{exp}(y))$ (for instance bird clutch size within species).

When you use simple Poisson glm on over-dispersed data you artificially decrease SE and p-value: you are going to find lots of false positive. When you use simple Poisson glm on under-dispersed data you artificially increase SE and p-value: you loose statistical power.

There are at least three options to obtain proper inference/prediction with count data:

* Use quasi-likelihood methods with approximate correction of uncertainty. That is available as glm(family="quasipoisson")
* Use multiplicative dispersal model such as negative-binomial.
* If the data are assumed to be over-dispersed conditional on predictors, use an "observation-level random effect"; that's equivalent to adding a residual to the Poisson model.


## Practice

1. Load the data reproduction.csv
```{r}
repro <- read.csv("data/reproduction.csv")

 # set.seed(1234)
 # repro$size <- 1+abs(1+ rnorm(n=nrow(repro), mean = 2, sd = 2))
 # repro$reproduction <- sapply(exp(0.2+0.3*(repro$size-4) + 0.2*repro$sex + rnorm(nrow(repro),0,sd=1.2)),
 #        function(x) rpois(n = 1, lambda = x))
 # 
 # ggplot(repro, aes(x=size, y=reproduction)) +
 #   geom_point()+ geom_smooth(method="lm")

write.csv(repro, file = "data/reproduction.csv", row.names = FALSE, quote = FALSE)
str(repro)
summary(repro)
```

2. Plot reproduction data, calculate the mean and variance in reproduction.
```{r}
mean(repro$reproduction)
var(repro$reproduction)

ggplot(repro, aes(x=reproduction)) + geom_density()
```

The variance is larger than the mean; data are currently over-dispersed (although it is possible that data are not over-dispersed conditional on predictors.)

3. Overlay a Gaussian distribution of same mean and variance, does it fit?
```{r}
normdens <- data.frame(x=seq(-10,max(repro$reproduction), by = 0.1), 
                       y=dnorm(x=seq(-10, max(repro$reproduction), by = 0.1),
                  mean = mean(repro$reproduction),
                  sd = sd(repro$reproduction)))

ggplot(repro, aes(x=reproduction)) + geom_histogram(aes(y=stat(count) / sum(count))) +
  geom_line(data=normdens, aes(x = x, y=y), inherit.aes = FALSE, color="red")

```

4. Fit an compare a lm and a Poisson glm of reproduction on size
```{r}
lmrepro <- lm(reproduction ~ size, data=repro)

glmrepro <- glm(reproduction ~ size, data = repro, family = "quasipoisson")

ggplot(repro, aes(x=size, y=reproduction)) + geom_point()+
  geom_smooth(method="lm",fullrange=TRUE, color="orange", fill="orange", alpha=0.2) +
  geom_smooth(method="glm", method.args=list(family="poisson"), color="red", fill="red", fullrange=TRUE, alpha=0.2) +
  geom_smooth(method="glm", method.args=list(family="quasipoisson"), color="blue", fill="blue", fullrange=TRUE, alpha=0.2) +
  ylim(c(-1,30)) + xlim(c(0,max(repro$size)))


```


5. Check the diagnostic plots for both models. Should you be worried?
```{r}
plot(lmrepro)
check_model(lmrepro)
plot(lmrepro)
check_model(glmrepro)
```

No worries. GLMs have different assumptions than lm(). Plots for glm generally look weird even if everything is good.

6. Extract and visualize a model prediction from both models (use the function
predict, and/or do it by hand to practice link-function back-transformation)

```{r}
sizex <- seq(from=0, to=10, by=0.1)
reprolatent <- coef(glmrepro)[1] + sizex*coef(glmrepro)[2]
#equivalent to:
predict(glmrepro, type = "link", newdata = data.frame(size=sizex))

pred1 <- data.frame(sizex=sizex, reprolatent=reprolatent)

ggplot(repro, aes(x=size, y=reproduction)) + geom_point() + 
  geom_line(data = pred1, aes(x=sizex, y=reprolatent), color="red")

pred1$reprodata <- exp(pred1$reprolatent)
#equivalent to:
predict(glmrepro, type = "response", newdata = data.frame(size=sizex))


ggplot(repro, aes(x=size, y=reproduction)) + geom_point() + 
  geom_line(data = pred1, aes(x=sizex, y=reprolatent), color="red")+
  geom_line(data = pred1, aes(x=sizex, y=reprodata), color="blue")

```


7. Before GLMs, researchers used to log-transform the data and fit linear models.
What are the problems with this approach?

```{r, eval=FALSE}
summary(lm(log(reproduction) ~ size, data=repro))

summary(lm(log(reproduction+1) ~ size, data=repro))

summary(lm(log(reproduction+0.1) ~ size, data=repro))

summary(lm(log(reproduction+0.001) ~ size, data=repro))

summary(lm(log(reproduction+0.000000000000000001) ~ size, data=repro))

#let's extract the slope parameter for smaller and smaller deviations from zero

sapply(X = 1:30, function(x){
  coef(lm(log(reproduction+10^(-x)) ~ size, data=repro))[2]
  }
  )

```

You cannot take the log of zero. The value of the slope is dependent on the arbitrary choice of a small deviation from zero.


### Factorial model

```{r}
glmreprosex <- glm(reproduction ~ sex, data = repro, family = "quasipoisson")
summary(glmreprosex)

exp(coef(glmreprosex)[2])

mean(repro$reproduction[repro$sex==1])/mean(repro$reproduction[repro$sex==0])


mean(repro$reproduction[repro$sex==0]) * exp(coef(glmreprosex)[2])
```

Does it work with the continuous scale of our previous model?


```{r}
predictedmeanat0 <- exp(coef(glmrepro)[1])
predictedmeanat1 <- exp(coef(glmrepro)[1] + coef(glmrepro)[2])

predictedmeanat1/predictedmeanat0
exp(coef(glmrepro)[2])
```


## Practice 2

```{r}
voles <- read.csv("data/voles.csv")
summary(voles)
```


## Over-dispersion as a random effect

```{r}
glmreprosex <- glm(reproduction ~ sex + size, data = repro, family = "quasipoisson")
summary(glmreprosex)

mreproWRONG <- glmmTMB(reproduction ~ sex +size, data = repro, family = "poisson")
summary(mreproWRONG)

repro$row <- 1:nrow(repro)
mrepro2 <- glmmTMB(reproduction ~ sex +size + (1|row), data = repro, family = "poisson")
summary(mrepro2)

#similar to :
mrepro3 <- glmmTMB(reproduction ~ sex +size , data = repro, family = "nbinom1")
summary(mrepro3)


mrepro4 <- glmmTMB(reproduction ~ sex +size + (1|row), data = repro, family = "nbinom2")
summary(mrepro4)
```

