---
title: "Mixed models 2"
author: "Timothee Bonnet"
date: "13 May 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Today we learn:**

* Interpreting and visualizing mixed-effect models
* Measuring the uncertainty in a random effect parameter
* (Should you treat "Year" as a fixed or random effect?)

Load packages for today:

```{r}
library(lmerTest)
library(ggplot2)
```


Load data:

```{r}
download.file("https://timotheenivalis.github.io/data/thorndata.csv", 
              destfile = "data/thorndata.csv")

download.file("https://timotheenivalis.github.io/data/droughtdata.csv", 
                        destfile = "data/droughtdata.csv")

download.file("https://timotheenivalis.github.io/data/roo.csv", 
                        destfile = "data/roo.csv")

download.file("https://timotheenivalis.github.io/data/storks.csv", 
                        destfile = "data/storks.csv")

```

## Visualize

We model the relationship EscapeDistance / Tail_Length. 
```{r}
roo <- read.csv("data/roo.csv")

ggplot(roo, aes(x=Tail_Length, y=EscapeDistance)) + 
  geom_point(aes(color=as.factor(Year)), alpha=0.5) + 
  geom_smooth(method = "lm")


mm0 <- lmer(EscapeDistance ~ Tail_Length + (1|Year), data = roo)
summary(mm0)

# Fixed effects:
fixef(mm0)

# Random deviations (a.k.a. "random predictions"):
ranef(mm0)
# Fixed effects plus random deviations:
coef(mm0)

#Random effect variance components
as.numeric(VarCorr(mm0)) 
```

Visualize the year slope predictions:

```{r}
predlm <- as.data.frame(coef( mm0 )$Year)
predlm$Year <- rownames(predlm)
mainslope <- as.data.frame(t(fixef(mm0)))

ggplot(roo, aes(x=Tail_Length, y=EscapeDistance)) + 
  geom_point(aes(color=as.factor(Year)), alpha=0.5) + 
  geom_smooth(method = "lm")+
  geom_abline(inherit.aes = FALSE, 
              data = predlm,
              aes(intercept=`(Intercept)`,
                  slope=Tail_Length, 
                  color=as.factor(Year))) + 
  geom_abline(inherit.aes = FALSE, 
              data = mainslope,
              aes(intercept=`(Intercept)`,
                  slope=Tail_Length), size=2) 
  


```

Visualize the distribution of year effects and where to expect new years:
```{r}
yeardeviations <- as.data.frame(ranef(mm0)$Year)
yeardeviations$Year <- row.names(yeardeviations)

ggplot(yeardeviations, aes(x=`(Intercept)`, color=Year)) + 
  geom_vline(aes(xintercept=`(Intercept)`, color=Year), size=2)

as.numeric(VarCorr(mm0))

xgrid <- seq(-8,8, by = 0.01)
randomdens <- dnorm(x = xgrid, mean = 0, sd = sqrt(as.numeric(VarCorr(mm0))))
REdensity <- data.frame(xgrid=xgrid, randomdens=randomdens)

ggplot(REdensity, aes(x=xgrid, y=randomdens)) + geom_line()+
  geom_vline(data= yeardeviations,
             aes(xintercept=`(Intercept)`, color=Year), size=2, alpha=0.5)
```

Visualize the projected range of observed and unobserved years around linear regression:
```{r}

Int_low95 <- fixef(mm0)["(Intercept)"] - 1.96 * sqrt(as.numeric(VarCorr(mm0)))
Int_upp95 <- fixef(mm0)["(Intercept)"] + 1.96 * sqrt(as.numeric(VarCorr(mm0))) 

Tail_Length <- seq(20,75, by = 1)


intervalRE <- data.frame(Tail_Length=Tail_Length,
                         low95 =  Int_low95 + fixef(mm0)["Tail_Length"]*Tail_Length,
                         upp95 = Int_upp95 + fixef(mm0)["Tail_Length"]*Tail_Length)

ggplot(roo, aes(x=Tail_Length, y=EscapeDistance)) + 
  geom_point(aes(color=as.factor(Year)), alpha=0.5) + 
  geom_abline(inherit.aes = FALSE, 
              data = predlm,
              aes(intercept=`(Intercept)`,
                  slope=Tail_Length, 
                  color=as.factor(Year))) + 
  geom_abline(inherit.aes = FALSE, 
              data = mainslope,
              aes(intercept=`(Intercept)`,
                  slope=Tail_Length), size=2) +
  geom_ribbon(inherit.aes = FALSE,
                data=intervalRE,
                aes(x=Tail_Length,ymin=low95, ymax=upp95), alpha=0.3)

# Adding residual variation for the year 2006:

Int_low95 <- coef(mm0)$Year["2013", "(Intercept)"] - 1.96 * sigma(mm0)
Int_upp95 <- coef(mm0)$Year["2013", "(Intercept)"] + 1.96 * sigma(mm0)

Tail_Length <- seq(20,75, by = 1)

intervalPrediction2013 <- data.frame(Tail_Length=Tail_Length,
                         low95 =  Int_low95 + fixef(mm0)["Tail_Length"]*Tail_Length,
                         upp95 = Int_upp95 + fixef(mm0)["Tail_Length"]*Tail_Length)

ggplot(roo, aes(x=Tail_Length, y=EscapeDistance)) + 
  geom_point(aes(color=as.factor(Year)), alpha=0.5) + 
  geom_abline(inherit.aes = FALSE, 
              data = predlm,
              aes(intercept=`(Intercept)`,
                  slope=Tail_Length, 
                  color=as.factor(Year))) + 
  geom_abline(inherit.aes = FALSE, 
              data = mainslope,
              aes(intercept=`(Intercept)`,
                  slope=Tail_Length), size=2) +
  geom_ribbon(inherit.aes = FALSE,
                data=intervalRE,
                aes(x=Tail_Length,ymin=low95, ymax=upp95), alpha=0.3)+
  geom_ribbon(inherit.aes = FALSE,
                data=intervalPrediction2013,
                aes(x=Tail_Length,ymin=low95, ymax=upp95, fill="2013"), alpha=0.3)
```


Repeat the process for id, a random effect that captures far more variation:
```{r}

mmid <- lmer(EscapeDistance ~ Tail_Length + (1|id), data = roo)
summary(mmid)


Int_low95 <- fixef(mmid)["(Intercept)"] - 1.96 * sqrt(as.numeric(VarCorr(mmid)))
Int_upp95 <- fixef(mmid)["(Intercept)"] + 1.96 * sqrt(as.numeric(VarCorr(mmid))) 

Tail_Length <- seq(20,75, by = 1)

intervalRE <- data.frame(Tail_Length=Tail_Length,
                         low95 =  Int_low95 + fixef(mmid)["Tail_Length"]*Tail_Length,
                         upp95 = Int_upp95 + fixef(mmid)["Tail_Length"]*Tail_Length)

mainslope <- as.data.frame(t(fixef(mmid)))

ggplot(roo, aes(x=Tail_Length, y=EscapeDistance)) + 
  geom_point(aes(color=as.factor(Year)), alpha=0.5) + 
  geom_abline(inherit.aes = FALSE, 
              data = mainslope,
              aes(intercept=`(Intercept)`,
                  slope=Tail_Length), size=2) +
  geom_ribbon(inherit.aes = FALSE,
                data=intervalRE,
                aes(x=Tail_Length,ymin=low95, ymax=upp95), alpha=0.3)
```



## Year as fixed or random effect?

```{r, echo=FALSE}
set.seed(123)
year <- 1980:2020
storks <- sapply(year-mean(year), function(x) rpois(n = 1, lambda = exp(0.2+0.1*x)))
devbabies <-rnorm(n = length(year), sd = 1)

method <- letters[1:10]

data <- expand.grid(year=year, method=method)
data$storks <-  sapply(data$year-mean(data$year), function(x) rpois(n = 1, lambda = exp(0.2+0.3*x))) + rep(1:length(method), each=length(year))
data$babies <- round(1000 + data$year*0.1+rnorm(n = length(data$year)) + rep(devbabies, length(method))) + rnorm(n = length(data$year))

keep <- as.logical(sapply(data$year-mean(year), function(x) rbinom(n = 1, size = 1, prob = plogis(0.1+0.3*x))))

data <- data[keep,]
nrow(data)
table(data$year)
write.csv(data, file = "data/storks.csv", row.names = FALSE, quote = FALSE)
```

Year as a fixed effect models a linear change in the response with years. This does not correct for the non-independence between data collected on a given year.
```{r}
summary(lm(babies ~ 1 + year, data = data))
summary(lm(babies ~ 1 + storks, data = data))
summary(lm(babies ~ 1 + storks +  year, data = data))
```

Year as a random effect models differences among years, without considering that years come in a specific order. The random effect correct the estimation of the fixed effect for non-independence between data collected on a give year. However, the random effect does not correct for linear changes in other fixed effects with time.
```{r}
summary(lmer(babies ~ 1 + (1|year), data = data))
summary(lmer(babies ~ 1 + storks + (1|year), data = data))
```

Sometimes you need both.
Year can be both a fixed and a random effect, meaning we want to model a linear change, and the dependence between data collected on the same years.
```{r}
summary(lmer(babies ~ 1 + storks +  year + (1|year) , data = data))
```


Try with the roo dataset:

```{r}
summary(lm(EscapeDistance ~ Tail_Length + Year , data = roo))
```

```{r}
summary(lmer(EscapeDistance ~Tail_Length + (1|Year), data = roo))

```

```{r}
summary(lmer(EscapeDistance ~Tail_Length + Year + (1|Year), data = roo))
```


## Uncertainty in random effects

Lme4 summary does not show any measure of uncertainty in random effect variances. How do we know whether a random effect is statistically significant, or what range of values are likely?

You can test a random effect using a Likelihood Ratio Test that compares a model with and without the random effect. **The models need to be nested, which means one model is a subset of the other one**.
```{r}
mm0 <- lmer(EscapeDistance ~ Age*Sex + (1|Year), data = roo)
m0 <- lm(EscapeDistance ~ Age*Sex, data = roo)
anova(mm0, m0)
```


```{r}
mm1 <- lmer(EscapeDistance ~ Age*Sex + (1|Year) + (1|id), data = roo)
mm2 <- lmer(EscapeDistance ~ Age*Sex + (1|Year), data = roo)
mm3 <- lmer(EscapeDistance ~ Age*Sex + (1|id), data = roo)
anova(mm1, mm2)
anova(mm1,mm3)
```

Estimate confidence intervals of standard deviation parameters:

```{r}
confint(mm1)
```
.sig01 is the standard deviation for the first random effect, .sig02 for the second random effect, .sigma for the residual standard deviation. 

What about uncertainty in the random level predictions produced by ranef() ?
You can't with lme4, and it is very difficult in general with maximum likelihood methods. You need to go Bayesian (for instance with MCMCglmm), but we won't cover that here.




<!-- ```{r} -->
<!-- # value for Adult, when Tail_Length=0 -->
<!-- coef(mInt1)[1] -->

<!-- # value for Adult when Tail_Length=30 or 50 -->
<!-- coef(mInt1)[1] + 30 * coef(mInt1)[2]  -->
<!-- coef(mInt1)[1] + 50 * coef(mInt1)[2]  -->

<!-- # value for Juvenile when Tail_Length=0 -->
<!-- coef(mInt1)[1] + coef(mInt1)[3]  -->

<!-- # value for Juvenile when Tail_Length=30 or 50 -->
<!-- coef(mInt1)[1] + coef(mInt1)[3] + 30*(coef(mInt1)[2] + coef(mInt1)[4])   -->
<!-- coef(mInt1)[1] + coef(mInt1)[3] + 50*(coef(mInt1)[2] + coef(mInt1)[4])   -->

<!-- ``` -->

<!-- ## Interaction between two continuous variables -->

<!-- ```{r} -->
<!-- ggplot(roo, aes(x=Tail_Length, y=EscapeDistance, color=Date)) +  -->
<!--   geom_point() + geom_smooth(method = "lm", data = roo [roo$Date > 250,], -->
<!--         aes(x=Tail_Length, y=EscapeDistance, color=Date), inherit.aes = FALSE, color="red") + -->
<!-- geom_smooth(method = "lm", data = roo [roo$Date > 200 & roo$Date<=250,], -->
<!--                    aes(x=Tail_Length, y=EscapeDistance, color=Date), inherit.aes = FALSE, color="orange") + -->
<!-- geom_smooth(method = "lm", data = roo [roo$Date<=200,], -->
<!--                    aes(x=Tail_Length, y=EscapeDistance, color=Date), inherit.aes = FALSE, , color="yellow") -->

<!-- mInt2 <- lm(EscapeDistance ~ Tail_Length*Date, data=roo) -->
<!-- summary(mInt2) -->
<!-- anova(mInt2) -->
<!-- ``` -->

<!-- ## elements of lme4 -->

<!-- ## Uncertainty and tests in lme4 -->


<!-- ## Gapminder exercises -->

<!-- ```{r} -->
<!-- # download.file(url = "https://timotheenivalis.github.io/data/gapminder_data.csv", -->
<!-- #               destfile = "gapminder.csv") -->
<!-- gapminder <- read.csv(file = "gapminder.csv") -->
<!-- str(gapminder) -->
<!-- ``` -->

<!-- Is gdpPercap influenced by lifeExp? On average for all continent?  -->
<!-- Is the relationship different among continents? -->





